% (This file is included by thesis.tex; you do not latex it by itself.)

\begin{abstract}

% The text of the abstract goes here.  If you need to use a \section
% command you will need to use \section*, \subsection*, etc. so that
% you don't get any numbering.  You probably won't be using any of
% these commands in the abstract anyway.

Terrorism though having been present since antiquity gained widespread adoption as a perceived method to accomplish a political goal, ideological or religious change in the latter part of the 20th century and into the 21st century. Being able to detect changes in of behaviour, particularly changes in intensity in terrorism is important for economic and societal reasons. Being able to detect changes in intensity is therefore an important task. This thesis investigates a number of methodologies capable of being able to detect changes in behaviour using online open source terrorist incident databases. A number of methodologies are evaluated specifically for modelling count time series data. These include count regression techniques, HMM's, time series aberration detection techniques (RAD and SURUS) and syndromic surveillance methods. The different methodologies are evaluated and the strengths and weaknesses of the different approaches are discussed. Using open source software the demonstration of the application of multiple techniques is illustrated and specifically the power of using multiple techniques in the diagnosis of changes in intensity of terrorism. 

\end{abstract}
